---
title: "Problem Set #2"
author: "Ian Bach"
date: "2024-08-30"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Question #1
1).(2 points) Consider a special linear regression model in which each observation has its own slope coefficient: $y_i = x_i\beta_i + \epsilon_i$, $x_i \neq 0$, $i = 1, \dots, n$. What is the least squares estimator of $\beta_4$?

- $y_i = x_i\beta_i + \epsilon_i$

- $y_i$ is the dependent variable

- $x_i$ is the independent variable

- $\beta_i$ is the slope coefficient specific to the $i$-th observation

- $\epsilon_i$ is the error term

To estimate $\beta_4$, focus on the fourth observation:
- $y_4 = x_4\beta_4 + \epsilon_4$

The least squares method seeks to minimize the sum of squared residuals. The residual for the 4th observation:

- $\text{Residual}_4 = y_4 - x_4\beta_4$

The sum of squared residuals:

- $\text{RSS}(\beta_4) = (y_4 - x_4\beta_4)^2$

To find the minimum, we take the derivative of $\text{RSS}(\beta_4)$ with respect to $\beta_4$, set it to zero, and simplify:

- $y_4 = x_4\beta_4$

The least squares estimate $\hat{\beta}_4 = \frac{y_4}{x_4}$

Question #2

(A) Find the derivative of the objective function with respect to $\beta$.

The derivative of a function gives us the slope, or how the function changes as $\beta$ changes. To minimize the function, we need to find where the slope is zero.

Sum of Squared Errors:

- $\frac{\partial}{\partial \beta} \left(\sum_{i=1}^{n} \left(y_i - x_i\beta\right)^2\right)$

- The derivative of $\left(y_i - x_i\beta\right)^2$ with respect to $\beta$ involves taking the derivative of the inner term $\left(y_i - x_i\beta\right)$ and then multiplying by the derivative of the square:

- $-2 \sum_{i=1}^{n}\left(y_i - x_i\beta\right)$

Penalty:

The penalty term $\lambda\beta^2$ penalizes large values of $\beta$.

- $\frac{\partial}{\partial \beta} \left(\lambda\beta^2\right) = 2\lambda\beta$

Combining the Derivatives:

-$\frac{\partial f(\beta)}{\partial \beta} = -2\sum_{i=1}^{n} x_i \left(y_i - x_i\beta\right) + 2\lambda\beta$

(B) Derive an explicit formula for $\hat{\beta}_{\text{ridge}}$.

Setting the Derivative to Zero:

- $\frac{\partial f(\beta)}{\partial \beta} = -2\sum_{i=1}^{n} x_i\left(y_i - x_i\beta\right) + 2\lambda\beta = 0$

Simplify:

Divide the equation by 2:

- $-\sum_{i=1}^{n} x_i\left(y_i - x_i\beta\right) + \lambda\beta = 0$

Rearrange the equation to collect all terms involving $\beta$ on one side:

- $\sum_{i=1}^{n} x_i y_i = \sum_{i=1}^{n} x_i^2\beta + \lambda\beta$

Factor out $\beta$ on the right side:

- $\sum_{i=1}^{n} x_i y_i = \beta\left(\sum_{i=1}^{n} x_i^2 + \lambda\right)$

Solve for $\beta$:

- $\hat{\beta}{\text{ridge}} = \frac{\sum{i=1}^{n} x_i y_i}{\sum_{i=1}^{n} x_i^2 + \lambda}$

(C) What happens if $\lambda \rightarrow 0$? What Happens if $\lambda \rightarrow \infty$?

- When $\lambda$ approaches zero, the penalty term $\lambda\beta^2$ becomes negligible. This means that the Ridge Regression estimator approaches the Ordinary Least Squares (OLS) estimator.

- When $\lambda \rightarrow \infty$, the Ridge Regression estimator approaches zero, meaning the model completely shrinks the coefficient and effectively ignores the input.

Question #3

Consider the following problem related to Ridge Regression.

Define the objective function for Ridge Regression:

Given:

The objective function to minimize is $f(\beta) = \sum_{i=1}^{n} (y_i - x_i \beta)^2 + \lambda \beta^2$, 

where:
$y_i$ is the dependent variable
$x_i$ is the independent variable
$\beta$ is the coefficient to be estimated
$\lambda$ is the regularization parameter

Define the gradient of the function:
The gradient of the objective function is given by:

$\nabla f(\beta) = 2 \sum_{i=1}^{n} (x_i \beta - y_i) x_i + 2 \lambda \beta$

Initialize parameters for implementing gradient descent:
The parameters to be initialized include:

$\lambda = 10$: Regularization parameter
$\beta_0 = 0$: Initial value of $\beta$
$\alpha = 0.001$: Learning rate
$\epsilon = 10^{-4}$: Convergence criterion

```{r Implement Gradient Descent in R}
# Data
X <- c(2, 4, 6, -2, -2, -6)
y <- c(2.5, 5.8, 3.7, -1.1, -3.7, -5)

# Parameters
lambda <- 10
beta <- 0
alpha <- 0.001
epsilon <- 10^-4

# To store beta values at each iteration
beta_history <- c(beta)

# Gradient descent loop
repeat {
  gradient <- 2 * sum(X * (X * beta - y)) + 2 * lambda * beta
  new_beta <- beta - alpha * gradient
  beta_history <- c(beta_history, new_beta)
  
  if (abs(new_beta - beta) < epsilon) {
    break
  }
  
  beta <- new_beta
}

# Ridge estimate
ridge_estimate <- beta

```

```{r Plot the Values of β_t }
# Plot the values of beta_t
plot(beta_history, type = "l", col = "blue", lwd = 2,
     xlab = "Iteration", ylab = expression(beta[t]),
     main = "Gradient Descent for Ridge Regression")

```


```{r Output the Ridge Estimate}
cat("Ridge estimate:", ridge_estimate, "\n")
```


Question #4

(5 points) This question involves the Auto dataset, which is included in the package ISLR2.

(a) Fit a multiple linear regression with mpg as the response and all other variables except name as predictors. Use the summary() function to print the results.

```{r}
library(ISLR2)

Test <- ISLR2::Auto

model <- lm(mpg ~ . - name, data = Test)
summary(model)

```

b) Is there a relationship between the predictors and the response?

Outputs:

F-Statistic: 252.4 (A high value suggests a strong relationship)
p-value associated with the F-statistic: < 2.2e-16 (Indicates that it is very unlikely that all predictors have no relationship with mpg)
The P-value is extremely low, so we reject the null hypothesis.
Conclusion: Yes, there is a relationship between the predictors and the response variable mpg.

(c) For which of the predictors can you reject the null hypothesis $H_0 : \beta_j = 0$?

By looking at the P-values from the summary:

Displacement: p-value = 0.00844
Weight: p-value = 2e-16
Year: p-value = 2e-16
Origin: p-value = 4.67e-07
Conclusion: The predictors (Displacement, Weight, Year, and Origin) have p-values less than 0.05, meaning we reject the null hypothesis for these predictors. They are showing statistically significant relationships with mpg.

(d) If I obtain a p-value of 0.03 for the null hypothesis $H_0 : \beta_j = 0$, is it correct to interpret it as: given our sample, $\beta_j = 0$ with probability 0.03? Explain.

Explanation: A p-value of 0.03 indicates that if the hypothesis $H_0 : \beta_j = 0$ is true, there is a 3% chance of observing a test statistic as extreme as the one observed. It does not mean that the probability of $\beta_j = 0$ is 0.03.

Conclusion: No, it is incorrect to interpret a p-value of 0.03 as the probability that the null hypothesis is true. The p-value represents the probability of obtaining a test statistic at least as extreme as the one observed, assuming that the null hypothesis $H_0 : \beta_j = 0$ is true.