---
title: 'Problem Set #4'
author: "Ian Bach"
date: "2024-09-13"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1.  (10 points) This question involves the Default dataset, which is included in the package ISLR2.

(a) Fit a logistic regression that uses income and balance to predict default (Model 1). Report the estimates. [Hint: it should include an intercept.]
(b) Suppose we classify an individual to the default category with a threshold probability of 0.5. Compute the LOOCV test error estimate for Model 1.
(c) Now consider a logistic regression that predicts default using income, balance, and a dummy variable for student (Model 2). Compute the LOOCV test error estimate for this model. Should we include a dummy variable for student?

1A.

$$
\text{logit}(P(\text{default}=1)) = \beta_0 + \beta_1 \times \text{income} + \beta_2 \times \text{balance}
$$

```{r P1A}
# Load the necessary package and data
library(ISLR2)

# Fit a logistic regression model (Model 1)
model1 <- glm(default ~ income + balance, data = Default, family = "binomial")

# View the model summary to see the estimates
summary(model1)


```

1B.

```{r P1B}
# Load the boot package
library(boot)

# Define a function to calculate the prediction error for LOOCV
loocv_error <- function(model, data) {
  # LOOCV using cv.glm from the boot package
  loocv_result <- cv.glm(data, model, K = nrow(data))
  return(loocv_result$delta[1])  # LOOCV estimate
}

# Calculate LOOCV error for Model 1
loocv_model1 <- loocv_error(model1, Default)
loocv_model1


```

1C.

$$
\text{logit}(P(\text{default}=1)) = \beta_0 + \beta_1 \times \text{income} + \beta_2 \times \text{balance} + \beta_3 \times \text{student}
$$

```{r P1C}
# Fit a logistic regression model (Model 2) with student dummy variable
model2 <- glm(default ~ income + balance + student, data = Default, family = "binomial")

# Calculate LOOCV error for Model 2
loocv_model2 <- loocv_error(model2, Default)
loocv_model2

```

2.  (10 points) This question uses the bootstrap to estimate the standard errors under Model 2 in Exercise 1.

(a) Write a function, boot.estimates, that takes as input the Default dataset and an index of the observations, and outputs the coefficient estimates of income, balance and student in a logistic regression (with an intercept).
(b) Generate 1000 bootstrapped samples and use boot.estimates to estimate the standard errors of the coefficients corresponding to income, balance and student.

2A.

```{r 2A}
# Define the boot.estimates function
boot.estimates <- function(data, index) {
  # Subset the data using the provided index
  boot_sample <- data[index, ]
  
  # Fit a logistic regression model using the bootstrapped data
  model <- glm(default ~ income + balance + student, data = boot_sample, family = "binomial")
  
  # Return the coefficients of income, balance, and student
  return(coef(model)[c("income", "balance", "studentYes")])  # studentYes corresponds to the dummy for student
}

```

2B.

```{r 2B}
# Load necessary library
library(boot)

# Define the bootstrap function to calculate estimates
boot.estimates <- function(data, index) {
  boot_sample <- data[index, ]
  model <- glm(default ~ income + balance + student, data = boot_sample, family = "binomial")
  return(coef(model)[c("income", "balance", "studentYes")])
}

# Perform the bootstrap with 1000 replications
set.seed(123)  # For reproducibility
boot_results <- boot(Default, boot.estimates, R = 1000)

# View the bootstrap results
boot_results

# Calculate the standard errors of the coefficients
boot_se <- apply(boot_results$t, 2, sd)
boot_se

```
