---
title: "Problem Set 6"
author: "Ian Bach"
date: "2024-09-25"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1A

-   Polynomial for $x \leq \xi$ $f_1(x)$

For $x \leq \xi$, the term $(x - \xi)^3 = 0$, so the function simplifies to:

$f_1(x) = \beta_0 + \beta_1 x + \beta_2 x^2 + \beta_3 x^3$

Thus, $f_1(x)$ is a cubic polynomial for $x \leq \xi$ with coefficients:

$a_1 = \beta_0$

$b_1 = \beta_1$

$c_1 = \beta_2$

$d_1 = \beta_3$

-   Polynomial for $x > \xi$ is $f_2(x)$:

$f_2(x) = \beta_0 + \beta_1 x + \beta_2 x^2 + \beta_3 x^3 + \beta_4 (x - \xi)^3$

The term $\beta_4 (x - \xi)^3$ is:

$\beta_4 (x - \xi)^3 = \beta_4 (x^3 - 3\xi x^2 + 3\xi^2 x - \xi^3)$

The function $f_2(x)$ becomes:

$$
f_2(x) = \beta_0 + \beta_1 x + \beta_2 x^2 + \beta_3 x^3 + \beta_4 (x^3 - 3\xi x^2 + 3\xi^2 x - \xi^3)
$$

$$
f_2(x) = \beta_0 + \beta_1 x + \beta_2 x^2 + (\beta_3 + \beta_4) x^3 - 3\beta_4 \xi x^2 + 3\beta_4 \xi^2 x - \beta_4 \xi^3
$$

This will be written as:

$$
f_2(x) = (\beta_0 - \beta_4 \xi^3) + (\beta_1 + 3\beta_4 \xi^2) x + (\beta_2 - 3\beta_4 \xi) x^2 + (\beta_3 + \beta_4) x^3
$$

So the coefficients of $f_2(x)$ are:

$$
a_2 = \beta_0 - \beta_4 \xi^3
$$

$$
b_2 = \beta_1 + 3 \beta_4 \xi^2
$$

$$
c_2 = \beta_2 - 3 \beta_4 \xi
$$

$$
d_2 = \beta_3 + \beta_4
$$

The cubic polynomial for $x > \xi$ is:

$$
f_2(x) = (\beta_0 - \beta_4 \xi^3) + (\beta_1 + 3\beta_4 \xi^2) x + (\beta_2 - 3\beta_4 \xi) x^2 + (\beta_3 + \beta_4) x^3
$$

## Question 1B

For continuity at $x = \xi$, the values of $f_1(x)$ and $f_2(x)$ must be equal at $x = \xi$. That is:

$$
f_1(\xi) = f_2(\xi)
$$

Evaluate $f_1(\xi)$

$f_1(\xi) = \beta_0 + \beta_1 \xi + \beta_2 \xi^2 + \beta_3 \xi^3$

Evaluate $f_2(\xi)$

Evaluate $f_2(\xi) = (\beta_0 - \beta_4 \xi^3) + (\beta_1 + 3 \beta_4 \xi^2) \xi + (\beta_2 - 3 \beta_4 \xi) \xi^2 + (\beta_3 + \beta_4) \xi^3$

$$
f_2(\xi) = \beta_0 + \beta_1 \xi + \beta_2 \xi^2 + \beta_3 \xi^3
$$

$$
f_2(\xi) = \beta_0 + \beta_1 \xi + \beta_2 \xi^2 + (\beta_3 + \beta_4) \xi^3
$$

The following terms cancel out: $\beta_4$ and $\xi_3 \xi^3 \xi_3$

$$
f_2(\xi) = \beta_0 + \beta_1 \xi + \beta_2 \xi^2 + \beta_3 \xi^3
$$

$$
f_1(\xi) = f_2(\xi)
$$

## Question 1C

First, compute the first derivative of $f_1(x)$:

$$
f_1'(x) = \beta_1 + 2 \beta_2 x + 3 \beta_3 x^2
$$

At $x = \xi$:

$$
f_1'(\xi) = \beta_1 + 2 \beta_2 \xi + 3 \beta_3 \xi^2
$$

The first derivative of $f_2(x)$ is:

$$
f_2'(x) = \beta_1 + 2 \beta_2 x + 3 \beta_3 x^2
$$

$$
f_2'(x) = \beta_1 + 3 \beta_4 \xi^2 + 2 (\beta_2 - 3 \beta_4 \xi) x + 3 (\beta_3 + \beta_4) x^2
$$

At $x = \xi$:

$$
f_2'(\xi) = \beta_1 + 3 \beta_4 \xi^2 + 2 (\beta_2 - 3 \beta_4 \xi) \xi + 3 (\beta_3 + \beta_4) \xi^2
$$

Simplify:

$$
f_2'(\xi) = \beta_1 + 3 \beta_4 \xi^2 + 2 \beta_2 \xi - 6 \beta_4 \xi^2 + 3 \beta_3 \xi^2 + 3 \beta_4 \xi^2
$$

After combining like terms:

$$
f_2'(\xi) = \beta_1 + 2 \beta_2 \xi + 3 \beta_3 \xi^2
$$

Thus:

$$
f_1'(\xi) = f_2'(\xi)
$$

This confirms that the first derivative is continuous at $x = \xi$.

## Question 1D

The second derivative of $f_1(x)$ is:

$$
f_1''(x) = 2 \beta_2 + 6 \beta_3 x
$$

At $x = \xi$:

$$
f_1''(\xi) = 2 \beta_2 + 6 \beta_3 \xi
$$

The second derivative of $f_2(x)$ is:

$$
f_2''(x) = 2 (\beta_2 - 3 \beta_4 \xi) + 6 (\beta_3 + \beta_4) x
$$

At $x = \xi$:

$$
f_2''(\xi) = 2 (\beta_2 - 3 \beta_4 \xi) + 6 (\beta_3 + \beta_4) \xi
$$

Simplifying:

$$
f_2''(\xi) = 2 \beta_2 - 6 \beta_4 \xi + 6 \beta_3 \xi + 6 \beta_4 \xi
$$

The $\beta_4$ terms cancel out, so:

$$
f_2''(\xi) = 2 \beta_2 + 6 \beta_3 \xi
$$

Thus:

$$
f_1''(\xi) = f_2''(\xi)
$$

This confirms that the second derivative is continuous at $x = \xi$.

## Question 2A

$$
\lambda = \infty, m = 0
$$


-   When $\lambda = \infty$, the penalty term dominates. In this case, since $m = 0$, the penalty applies directly to $g(x)$ itself, implying that $g_b(x)$ must be constant. This is because the only way to minimize the penalty term $\int [g(x)]^2 dx$ when $\lambda$ is very large is for $g(x)$ to be a constant function.

-   **Result:** $g_b(x)$ is a constant function, likely the mean of the $y_i$'s.

## Question 2B

$$
\lambda = \infty, m = 1
$$


-   With $\lambda = \infty$ and $m = 1$, the penalty applies to the first derivative $g'(x)$. To minimize the penalty, $g'(x) = 0$, meaning $g(x)$ must be a constant function again.

-   **Result:** $g_b(x)$ is constant, same as in part (a).

## Question 2C

$$
\lambda = \infty, m = 2
$$


-   For $\lambda = \infty$ and $m = 2$, the penalty applies to the second derivative $g''(x)$. Minimizing the penalty forces $g''(x) = 0$, meaning $g_b(x)$ must be a linear function (since a function with zero second derivative is a straight line).

-   **Result:** $g_b(x)$ is a linear function, $g_b(x) = ax + b$, where $a$ and $b$ are constants.

## Question 2D

$$
\lambda = \infty, m = 3
$$


-   For $\lambda = \infty$ and $m = 3$, the penalty applies to the third derivative $g^{(3)}(x)$. Minimizing the penalty forces $g^{(3)}(x) = 0$, meaning $g_b(x)$ must be a quadratic function (since a function with zero third derivative is a quadratic polynomial).

-   **Result:** $g_b(x)$ is a quadratic function, $g_b(x) = ax^2 + bx + c$, where $a$, $b$, and $c$ are constants.

## Question 2E

$$
\lambda = \infty, m = 2
$$


-   When $\lambda = 0$, there is no penalty for roughness, so $g_b(x)$ will minimize the first term in the loss function alone. This would make $g_b(x)$ exactly interpolate the data points, meaning it passes through every $(x_i, y_i)$ pair. Since $m = 2$, we would expect a smooth curve, but the penalty term doesn't influence the result when $\lambda = 0$.

-   **Result:** $g_b(x)$ will interpolate the data points and can be any flexible, non-penalized smooth curve.

## Question 3A

```{r}
library(ISLR)
library(splines)

# Load the Wage dataset
data(Wage)

# Fit the model using cubic splines with knots at 30, 50, and 60
fit_cubic_spline <- lm(wage ~ bs(age, knots = c(30, 50, 60), degree = 3), data = Wage)

# Plot the data
plot(Wage$age, Wage$wage, col = "gray", xlab = "Age", ylab = "Wage")

# Add the fitted spline curve to the plot
age_grid <- seq(min(Wage$age), max(Wage$age), length.out = 100)
wage_preds <- predict(fit_cubic_spline, newdata = list(age = age_grid))
lines(age_grid, wage_preds, col = "blue", lwd = 2)

predict(fit_cubic_spline, newdata = list(age = 30))

```

## Question 3B

```{r}
# Fit the smoothing spline with cross-validation
fit_smoothing_spline <- smooth.spline(Wage$age, Wage$wage, cv = TRUE)

# Plot the data
plot(Wage$age, Wage$wage, col = "gray", xlab = "Age", ylab = "Wage")

# Add the fitted smoothing spline curve
lines(fit_smoothing_spline, col = "red", lwd = 2)

predict(fit_smoothing_spline, x = 30)$y

```

## Question 3C

```{r}
library(gam)

# Fit the GAM model
fit_gam <- gam(wage ~ ns(age, df = 4) + maritl, data = Wage)

# Plot the effects of the GAM model
plot(fit_gam, se = TRUE)

# Plot the effects of the GAM model
plot(fit_gam, se = TRUE)

predict(fit_gam, newdata = data.frame(age = 30, maritl = "2. Married"))

```
